# Assignment 6: Data Analysis excercise


```{r}
date()
```

## RATS

### Reading data into R
```{r}
# Load required libraries
library(tidyverse)
library(lme4)
library(gridExtra)

# Read the data into R 
RATSL <- read.table("/Users/minmaunu/IODS-project/data/RATSL.txt", header = TRUE)

# Change the categorical variables to factors
RATSL$ID <- as.factor(RATSL$ID)
RATSL$Group <- as.factor(RATSL$Group)

# Take a glimpse on the data
glimpse(RATSL)
```
Data has 176 rows and 5 columns. Everything seems to be in order and data is ready for the analysis.  

## Plot non-standardized data

Let's draw a line plot of the rat weights in diet groups 1-3

```{r}
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```

High start weight means high end weight, so the data should be standardized.

## Plot standardized data

```{r}
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdweight = (Weight-mean(Weight))/sd(Weight)) %>%
  ungroup()

ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$stdweight), max(RATSL$stdweight)), name = "standardized bprs")
```

Looks better. Next let's look at the response to the diets. 

## Response to the diets

```{r}
n <- RATSL$Time %>% unique() %>% length()
RATS_trmt <- RATSL %>% group_by(Group, Time) %>% summarise(mean = mean(Weight), se = sd(Weight)/sqrt(n))
glimpse(RATS_trmt)
```

```{r}
ggplot(RATSL, aes(x= as.factor(Time), y=Weight, fill=Group)) +
  geom_boxplot() +
  theme(legend.position = c(0.8,0.4), panel.grid = element_blank(), panel.background = element_blank(), panel.border = element_rect(fill = FALSE), legend.key = element_rect(fill=FALSE)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)") + 
  scale_x_discrete(name = "Time") + 
  scale_fill_manual(values=c("pink", "forestgreen", "lightblue")) +
  ggtitle("Rat weight over time according to different diets ")
```

On average diet 1 results lower weight compared to diet 2 and 3 in rats. It seems that diet 2 has the biggest variation on rats. 

Next let's look at the outlier of the data 

## Outlier in the data

```{r}
RATSL8 <- RATSL %>% 
  group_by(Group, ID) %>%
  summarize(mean = mean(Weight)) %>%
  ungroup()

ggplot(RATSL8, aes(x=Group, y=mean)) + 
  geom_boxplot() + 
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") + 
  theme(legend.position = c(0.8,0.8), panel.grid = element_blank(), panel.background = element_blank(), panel.border = element_rect(fill = FALSE)) +
  scale_y_continuous(name = "mean(Weight) per group")
```

One outlier can be found from each group, but in group 2 it clearly differs from the mean. Let’s remove it and draw the boxplot again!

```{r}
RATSL8S1 <- filter(RATSL8, (Group=="2" & mean<500) | Group!="2")

ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  theme(legend.position = c(0.8,0.8), panel.grid = element_blank(), panel.background = element_blank(), panel.border = element_rect(fill = FALSE)) +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight) per group")
```


## BPRSL

### Reading data into R

```{r}
# Read the data 
BPRSL <- read.table("/Users/minmaunu/IODS-project/data/BPRSL.txt", header = TRUE)

# Change the categorical variables to factors
BPRSL$treatment <- as.factor(BPRSL$treatment)
BPRSL$subject <- as.factor(BPRSL$subject)

# Take a glimpse on the data
glimpse(BPRSL)
```

Data has 360 rows and 5 columns. Everything seems to be in order and data is ready for the analysis. 

## Plot the data with a line plot
```{r}
ggplot(BPRSL, aes(x = week, y = bprs, fill=subject, color = treatment)) +
  geom_line(aes(linetype = treatment)) +
  theme(legend.position = c(0.8,0.8), panel.grid = element_blank(), panel.background = element_blank(),
        panel.border = element_rect(fill = FALSE), legend.key = element_rect(fill=FALSE)) + 
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 2), expand=c(0,0)) + 
  scale_y_continuous(name = "bprss") +
  scale_color_manual(values=c("pink", "forestgreen")) +
  ggtitle("Symptoms in patients over time")
```

It's quite busy plot, but just by eyeballing it seems that with the treatment 1 the symptoms have slightly decreased within time, whereas with treatment 2 there is more variation towards the end.

Next let’s fit a Random Intercept Model.

## Random Intercept Model

```{r}
# Load required libraries
library(Matrix)
library(lme4)
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

summary(BPRS_ref)
```


Now let’s fit a random intercept and random slope model and compare it to the random intercept model.

```{r}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

summary(BPRS_ref1)
```


```{r}
anova(BPRS_ref, BPRS_ref1)
```

Let’s try interaction between week and treatment and compare again.

```{r}
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)

summary(BPRS_ref2)
```


```{r}
anova(BPRS_ref2, BPRS_ref1)
```

Based on the chi-squared statistics and p-value of the likelihood ratio test, random intercept and random slope model is the best. Let’s plot:

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(gridExtra)
p1 <- ggplot(BPRSL, aes(x = week, y = bprs, fill = subject, color = treatment)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 2)) +
  scale_y_continuous(name = "Observed bprs") +
    scale_color_manual(values=c("pink", "forestgreen")) +
    ggtitle("Symptoms in patients over time") +
  theme(panel.grid = element_blank(), panel.background = element_blank(),
        panel.border = element_rect(fill = FALSE), legend.key = element_rect(fill=FALSE), legend.position = "bottom",)
Fitted <- fitted(BPRS_ref1)
BPRSL <- mutate(BPRSL, fitted=Fitted)

p2 <- ggplot(BPRSL, aes(x = week, y = fitted, fill = subject, color = treatment)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 2)) +
  scale_y_continuous(name = "Fitted bprs") +
  scale_color_manual(values=c("pink", "forestgreen")) +
    ggtitle(" ") +
  theme(panel.grid = element_blank(), panel.background = element_blank(),
        panel.border = element_rect(fill = FALSE), legend.key = element_rect(fill=FALSE), legend.position = "bottom")


# p1 and p2 side by side
grid.arrange(p1, p2, ncol=2)
```

Based on the fitted model, we can see thath the treatments are after all equally effective in reducing the symptoms.

Done!
